# Week 6 - Harvesting and bifurcation {-}

In this lab section, we're going to analyze the budworm population
dynamic model from [**Ludwig et al.,
1978**](https://doi.org/10.2307/3939).

<br> **Part 1 - Stability of the budworm model**

In part 1 we're going to visualize the stability of the budworm model,
by plotting the differential equation. We will plot the the differential
equation with different initial, which we will see that the number and stability of equilibrium changes when parameter changes.

\[
\dfrac{dN}{dt} = rN(1 - \dfrac{N}{K}) - \dfrac{HN^2}{A^2 + N^2}
\]

```{r}
#### Plotting the functional form for different parameters
#### Parameter setting
r <- 0.055; K <- 10; H <- 0.1; A <- 1

#### Visualize the whole dN/dt with different H
N.vec <- seq(from = 0, to = 10, length = 500)
H.breaks <- c(0.05, 0.12, 0.20)
dat <- outer(X = N.vec, Y = H.breaks,
             function(N, H){r * N * (1 - N / K) - (H * N^2 / (A^2 + N^2))})
matplot(x = N.vec, y = dat, type = "l", 
        xlab = "N", ylab = "dN/dt", col = "blue", lwd = 2, las = 1)
abline(h = 0)
legend("bottomleft", legend = H.breaks, title = "H", col = "blue", lty=1:3, lwd = 2)
```
<br>

Second, we're going to plot $harvest$ rate against $N$ with
separate components of $dN/Ndt$, which the blue line is 
\[
\dfrac{HN}{A^2 + N^2}
\]
with different $H$, the red line is, 
\[
r(1 - \dfrac{N}{K})
\]
and the points that blue line and red line crosses are the equilibrium points.

```{r}
#### Visualize separate components of dN/Ndt with different H
N.vec <- seq(from = 0, to = 10, length = 500)
H.breaks <- c(0.05, 0.12, 0.20)
dat.growth <- outer(X = N.vec, Y = H.breaks,
                    function(N, H){H * N / (A^2 + N^2)}) # Note notation change
matplot(x = N.vec, y = dat.growth, type = "l", ylim = c(0, 0.10), las = 1, 
        xlab = "N", ylab = "harvest rate", col = "blue", lwd = 2)
curve(r * (1 - x/K), add = T, col = "red", lwd = 2) # Just curve since its the same line, and note variable notation change
abline(h = 0)
legend("topright", legend = H.breaks, title = "H", col = "blue", lty=1:3, lwd = 2)
```
<br>


**Part 2 - Use `rootSolve` function `gradient` and `uniroot.all`, to solve stability of budworm model**

```{r}
#### Stability of the budworm model, as a function of its parameters
#### Using "rootSolve" function "gradient" and "uniroot.all"
#### Works best for simple models and those with known solutions
########################################################################################################################
library(rootSolve)
#### Parameter setting
r <- 0.055; K <- 10; H <- 0.1; A <- 1

#### Spruce budworm model for rootSolve
Budworm <- function(N, H = 0.1){
  r * N * (1 - N / K) - (H * N^2 / (A^2 + N^2))
}

#### Function of root stability
Stability <- function(H.value = 0.1){
  equilibrium <- uniroot.all(f = Budworm, interval = c(0, K), H = H.value) # finds all roots
  lambda <- vector(mode = "numeric", length = length(equilibrium))
  for(i in 1:length(equilibrium)){
    lambda[i] <- sign(gradient(f = Budworm, x = equilibrium[i], H = H.value))
  }
  return(list(Equilibrium = equilibrium,
              Lambda = lambda))
}

#### Bifurcation diagram for H
H.vec <- seq(0.001, 0.3, by = 0.0001)

## Create plotting frame
plot(0, xlim = range(H.vec), ylim = c(0, 10), type = "n", las = 1,
     xlab = "H", ylab = "Equilibrium density, N*", main = "Budworm model bifurcation along H")
legend("topright", pch = 15, pt.cex = 2, c("stable", "unstable"),
       col = c("darkblue", "lightblue"))

## Calculate number of roots and stability across range of H
for(H in H.vec){
  temp <- Stability(H.value = H)
  points(x = rep(H, length(temp$Equilibrium)),
         y = temp$Equilibrium,
         pch = 15, col = ifelse(temp$Lambda == -1, "darkblue", "lightblue"))
}
```
<br>

Take a look a this
[**website**](https://www.math.ksu.edu/~albin/teaching/math340_fall2018/slides/03_bifurcations.html) if you're interested in more details of bifurcation. 
<br>
<br>

**Extra materials**
<br>
**Using `deSolve` function `ode` to brute-force stable solution**

Here we're going to use `deSolve` to solve the budworm model,

```{r}
#### Budworm model for deSolve

library(deSolve)

BudwormODE <- function(times, state, parms) {
  with(as.list(c(state, parms)), {
    dN_dt = r * N * (1 - N / K) - (H * N^2 / (A^2 + N^2))
    return(list(c(dN_dt)))
  })
}

### Parameters setting
times <- seq(0, 5000, by = 100)
state <- c(N = 10)

#### Bifurcation diagram for H -- the forward branch
## Set first forward simulation and saving space
H.vec.forward <- seq(0.001, 0.25, by = 0.001)
parms <- c(r = 0.055, K = 10, H = H.vec.forward[1], A = 1)
temp <- ode(func = BudwormODE, times = times, y = state, parms = parms)
forward <- data.frame(H = H.vec.forward,
                      N = rep(temp[length(times), 2], length(H.vec.forward)))
## Run across forward vector, using previous step equilibrium as new initial state
for(i in 2:length(H.vec.forward)){
  state <- c(N = forward[i-1, 2] + 0.001) ## Remember to add small perturbation
  parms <- c(r = 0.055, K = 10, H = forward[i, 1], A = 1)
  temp <- ode(func = BudwormODE, times = times, y = state, parms = parms)
  forward[i, 2] = temp[length(times), 2]
}
#### Bifurcation diagram for H -- the backward branch
## Set first backward simulation and saving space
H.vec.backward <- rev(H.vec.forward)
parms <- c(r = 0.055, K = 10, H = H.vec.backward[1], A = 1)
temp <- ode(func = BudwormODE, times = times, y = state, parms = parms)
backward <- data.frame(H = H.vec.backward,
                       N = rep(temp[length(times), 2], length(H.vec.backward)))
## Run across backward vector, using previous step equilibrium as new initial state
for(i in 2:length(H.vec.backward)){
  state <- c(N = backward[i-1, 2] + 0.001)  ## Remember to add small perturbation
  parms <- c(r = 0.055, K = 10, H = backward[i, 1], A = 1)
  temp <- ode(func = BudwormODE, times = times, y = state, parms = parms)
  backward[i, 2] = temp[length(times), 2]
}

#### Plot both forward and backward branch
plot(forward[, 1], forward[, 2],
     xlim = range(H.vec.forward), ylim = c(0, 10), las = 1, pch = 1, col = "darkblue", cex = 1.6,
     xlab = "H", ylab = "Equilibrium density, N*", main = "Budworm model bifurcation along H")
points(backward[, 1], backward[, 2], pch = 16, col = "lightblue")
legend("topright", pch = c(1, 16), pt.cex = 1.5, c("forward", "backward"),
       col = c("darkblue", "lightblue"))
```
<br>

<!-- **Part 2 - Non-dimensionalized model, a 2-D bifurcation diagram** -->
<!-- <br> -->

<!-- Third, we will plot a 2-D bifurcation diagram, with $r$ against $K$,  -->
<!-- \[ -->
<!-- \dfrac{dN}{dt} = rx(1 - \dfrac{x}{K}) - \dfrac{x^2}{1 + x^2}, -->
<!-- \] -->

<!-- \[ -->
<!-- r = \dfrac{2x^3}{(1 + x^2)^2}, K = \dfrac{2x^3}{(x^2 - 1)} -->
<!-- \] -->

<!-- ```{r} -->
<!-- #### 2-D bifurcation diagram following analytical solution from Strogatz -->
<!-- #### For nondimensionalized model dx/du = rx(1-x/k) - x^2/(1+x^2) -->
<!-- #### There exist parametric curves for bifurcation points r(x) and k(x) -->
<!-- x.bifur = seq(1.001, 50, by = 0.001) -->
<!-- bifur2D = data.frame(x.bifur = x.bifur,  -->
<!--                      r.bifur = 2 * x.bifur^3 / ((1 + x.bifur^2)^2), -->
<!--                      k.bifur = 2 * x.bifur^3 / (x.bifur^2 - 1)) -->
<!-- plot(r.bifur ~ k.bifur, bifur2D, type = "l", lwd = 2, las = 1,  -->
<!--      xlim = c(0, 60), ylim = c(0, max(bifur2D$r.bifur) * 1.1),  -->
<!--      xlab = "K", ylab = "r", main = "Budworm model bifurcation along r-k space") -->
<!-- ``` -->
<!-- <br> -->
