# Week 13 - Parameter space for apparent competition model {-} 

Here, we are going to visualize the outcomes of apparent competition in the $r_1/a_1$ -- $r_2/a_2$ space:
\begin{align*}
\frac{dN_1}{dt} &= r_1N_1(1-\frac{N_1}{K_1}) - a_1N_1P\\
\frac{dN_2}{dt} &= r_2N_2(1-\frac{N_2}{K_2}) - a_2N_2P\\
\frac{dP}{dt} &= e_1a_1N_1P + e_2a_2N_2P - mP.
\end{align*}
Two prey $N_1$ and $N_2$ grow logistically with intrinsic growth rate $r_1$ and $r_2$ and carrying capacity $K_1$ and $K_2$. Predator $P$ consumes two prey following Holling type I functional response, with $a_1$ and $a_2$ representing the consumption rate of $N_1$ and $N_2$ by predator respectively. After consumption, the assimilation rate are $e_1$ and $e_2$. The death rate of predator is represented by $m$.

**Part 1: Directly simulate dynamical system**

```{r}
library(deSolve)
library(ggplot2)

#### ODE setup
ApparentCompetition <- function(times, state, parms) {
  with(as.list(c(state, parms)), {
    dN1_dt = r1 * N1 * (1 - N1 / K1) - a1 * N1 * P
    dN2_dt = r2 * N2 * (1 - N2 / K2) - a2 * N2 * P
    dP_dt = e1 * a1 * N1 * P + e2 * a2 * N2 * P - d * P
    return(list(c(dN1_dt, dN2_dt, dP_dt)))
  })
}

times <- seq(0, 3000, by = 10)
state <- c(N1 = 0.1, N2 = 0.1, P = 1)
parms <- c(r1 = 0.1, r2 = 0.1, a1 = 0.35, a2 = 0.35, e1 = 0.4, e2 = 0.5, d = 0.8, K1 = 15, K2 = 12) # Omitting r1 and r2
 

#### Set up parameter space of interest
r_min <- 0.1
r_max <- 0.7
n <- 100
r1_vec <- seq(r_min, r_max, length.out = n)
r2_vec <- seq(r_min, r_max, length.out = n)
comp_out <- expand.grid(r1 = r1_vec, r2 = r2_vec)


#### Create saving space for simulation output
comp_out <- as.data.frame(cbind(comp_out,
                                matrix(0, 
                                       nrow = dim(comp_out)[1], 
                                       ncol = length(state))))
names(comp_out) <- c("r1", "r2", "N1", "N2", "P")


#### Simulate the ODE across the parameter space
start_time <- Sys.time()
for(i in 1:dim(comp_out)[1]){
  
  temp_parms <- parms
  temp_parms["r1"] <- comp_out[i, ]$r1
  temp_parms["r2"] <- comp_out[i, ]$r2
  
  temp_out <- ode(func = ApparentCompetition, 
                  times = times, 
                  y = state, 
                  parms = temp_parms)
  
  N_final <- nrow(na.omit(temp_out)) 
  comp_out[i, 3:5] <- temp_out[N_final, -1]
  
}
end_time <- Sys.time()
end_time - start_time

```

```{r}

#### Determine outcome status
extinct_thres <- 1e-7
comp_out$Outcome <- 
  ifelse(comp_out[, 3] < extinct_thres, "N2 win", 
         ifelse(comp_out[, 4] < extinct_thres, "N1 win", "Coexist"))


#### Visualize the competition outcomes in the grids
ggplot(comp_out, aes(x = r1/parms[3], y = r2/parms[4], z = Outcome, fill = Outcome)) +
  geom_tile() +
  geom_abline(intercept = 0, slope = (1 - parms[7] / (parms[5] * parms[3] * parms[8])), linewidth = 1.5) +
  geom_abline(intercept = 0, slope = (1 - parms[7] / (parms[6] * parms[4] * parms[9]))^-1, linewidth = 1.5) +
  labs(x = expression(r[1]/a[1]), y = expression(r[2]/a[2])) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_brewer(palette = "Set1") +
  theme_bw(base_size = 14) +
  theme(panel.grid = element_blank()) +
  coord_fixed(ratio = 1) 


```

**Part 2: Use analytical calculations**

```{r}

#### Create saving space for simulation output
r_min <- 0.1 # also try 0.01
r_max <- 0.7 # also try 1.75
n <- 100 # also try 200
r1_vec <- seq(r_min, r_max, length.out = n)
r2_vec <- seq(r_min, r_max, length.out = n)
comp_out_igr <- expand.grid(r1 = r1_vec, r2 = r2_vec)
comp_out_igr <- as.data.frame(cbind(comp_out_igr,
                                    matrix(0,
                                           nrow = dim(comp_out_igr)[1], 
                                           ncol = 2)))
names(comp_out_igr) <- c("r1", "r2", "IGR1", "IGR2")


#### Calculate the IGR across the parameter space
start_time <- Sys.time()
for(i in 1:dim(comp_out_igr)[1]){
  
  temp_parms <- parms
  temp_parms["r1"] <- comp_out_igr[i, ]$r1
  temp_parms["r2"] <- comp_out_igr[i, ]$r2
  
  comp_out_igr[i, 3:4] <- 
    with(as.list(temp_parms), {
      igr_1 <- r1 - a1 * ((r2/a2) * (1 - (d/(e2*a2*K2))))
      igr_2 <- r2 - a2 * ((r1/a1) * (1 - (d/(e1*a1*K1))))
      c(igr_1, igr_2) 
    })

}
end_time <- Sys.time()
end_time - start_time
```

```{r}

#### Determine outcome status
comp_out_igr$Outcome <- 
  ifelse(comp_out_igr[, 3] < 0, "N2 win", 
          ifelse(comp_out_igr[, 4] < 0, "N1 win", "Coexist"))


#### Visualize the competition outcomes in the grids
ggplot(comp_out_igr, aes(x = r1/parms[3], y = r2/parms[4], z = Outcome, fill = Outcome)) +
  geom_tile() +
  geom_abline(intercept = 0, slope = (1 - parms[7] / (parms[5] * parms[3] * parms[8])), linewidth = 1.5) +
  geom_abline(intercept = 0, slope = (1 - parms[7] / (parms[6] * parms[4] * parms[9]))^-1, linewidth = 1.5) +
  labs(x = expression(r[1]/a[1]), y = expression(r[2]/a[2])) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_brewer(palette = "Set1") +
  theme_bw(base_size = 14) +
  theme(panel.grid = element_blank()) +
  coord_fixed(ratio = 1)

```


